The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile

### 2024-04-04 article/paper_drought_Chile2023.qmd
table landcover

---
title: The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile
#subtitle: A Short Subtitle
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.cl
    affiliations: 
        - id: umayor
          name: Universidad Mayor
          department: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad
          #address: Street Address
          city: Santiago, Chile
          #state: State
          postal-code: 7500994
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Anton Vrieling
    email: a.vrieling@utwente.nl
    affiliations:
        - id: itc
          name: University of Twente
          department: Faculty of Geo-Information Science and Earth Observation (ITC)
          #address: Street Address
          city:  Enschede, The Netherlands
  - name: Francisco Meza
  - name: Iongel Duran-Llacer
  - name: Francisco Fernández
    email: francisco.fernandez@uss.cl
    affiliations:
        - id: uss
          name: Universidad San Sebastian
          department: 
          #address: Street Address
          city:  
  - name: Alejandro Venegas-González
  - name: Nicolas Raab
  - name: Dylan Craven        
  #   note: |
  #     Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  # - name: Cat Memes
  #   email: cat@example.com
  #   affiliations:
  #       - ref: another-u
  #   note: Yet another author footnote.
  # - name: Derek Zoolander
  #   email: derek@example.com
  #   affilations:
  #       - ref: some-tech
abstract: |
 The north-central region of Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a megadrought. The aim of our study is to evaluate the land cover change over continental Chile and to examine how this is connected to drought indices of water supply, atmospheric evaporative demand (AED), soil moisture, and their effects on vegetation productivity. The drought indices were derived using monthly ERA5-Land reanalysis data spanning from 1981 to 2023. The Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets were utilized to obtain information on annual land cover and monthly vegetation productivity. We analyzed short- (1, 3, 6 months) to long-term (12, 24, 36 months) time scales of drought. Our results showed that land cover change was highest in the south-central part of the country, reaching changes as high as 36% in the surface type. The water demand has increased for the whole country, with a major increase in the north. The AED and soil moisture evidence a decreasing trend, which decreases at longer time scales and from north to south. The extreme south part of the country shows an increase in supply. Vegetation productivity has a negative trend in the north-central region for all land cover types. On the other hand, forests seem to be the most resistant type to drought. The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands. The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are the ones that consider soil moisture for twelve months and the combined effect of precipitation and AED for 24 and 12 months. The results indicate that the north-central region is the most sensitive to water supply deficits lasting longer than a year. 
keywords: 
  - drought
  - land cover change
  - vegetation productivity
  - satellite
date: last-modified
bibliography: references.bib
keep-tex: false
include-before-body:
  - text: |
      \captionsetup{justification=raggedright,singlelinecheck=false}
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      layout: onecolumn
      model: 3p
      cite-style: authoryear
      natbiboptions: numbers
    include-in-header: 
      text: \usepackage{lineno}\linenumbers
            \usepackage{multirow}
            \usepackage{lscape}
            \newcommand{\blandscape}{\begin{landscape}}
            \newcommand{\elandscape}{\end{landscape}}
  elsevier-docx:
    keep-tex: true
    number-sections: true
    journal:
      name: Journal Name
      formatting: preprint
      layout: onecolumn
      model: 3p
      cite-style: authoryear
      natbiboptions: numbers
    include-in-header: 
      text: \usepackage{lineno}\linenumbers
            \usepackage{multirow}
            \usepackage{lscape}
            \newcommand{\blandscape}{\begin{landscape}}
            \newcommand{\elandscape}{\end{landscape}}
---

@@

For the purpose of monitoring water supply drought, we used the well-known Standardized Precipitation Index (SPI), which relies on precipitation data. To evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI), developed by @Hobbins2016 and @McEvoy2016, which is based on the AED. The United States currently monitors drought using the EDDI (https://psl.noaa.gov/eddi/) as an experimental index. To consider the combined effect of water supply and demand, we selected the SPEI [@Vicente-Serrano2010]. For SPEI, an auxiliary variable $D = P-AED$ is calculated. Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions [@Li2022]. Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) [@Hao2013;@AghaKouchak2014]. In our case, for the SSI, we used the average soil moisture from ERA5L at 1m depth. Finally, for the proxy of productivity, we used the zcNDVI proposed by @Zambrano2018, which was derived from the monthly time series of NDVI retrieved from MOD13A1. All the indices are multi-scalar and can be used for the analysis of short- to long-term droughts.

@@

The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 1981–2023 in order to be used for short- to long-term evaluation of drought. In the case of the proxy of vegetation productivity (zcNDVI), it was calculated for a time scale of six months at monthly frequency for 2000–2023. For zcNDVI, we test time scales of 1, 3, 6, and 12 months in December and its correlation with net primary production (NPP) obtained from the MOD17A3HGF product from MODIS. We choose to use six months because r-squared with NPP increases from one to six months and from six to 12 months has little improvement (see supplementary material in Section S5).

@@

## Future outlook (to complete)

@@

There is a trend toward decreasing water supply in most parts of Chile, particularly in the “Centro” and “Norte Chico” regions. The whole country showed an increase in AED. Vegetation productivity only showed a decrease in the “Norte Chico” and “Centro,” being highest for shrubland and croplands. Forest is the land cover most resistant to drought, as shown along Chile, and shrubland and cropland are the most sensitive.

A soil moisture deficit of 12 months (SSI-12) is highly correlated with vegetation productivity for the land cover classes of shrubland, savannas, croplands, and forest in “Norte Chico” and “Centro.” For the southern part of the country with humid conditions, the correlation with SSI decreases. Soil moisture overcomes the capacity to explain vegetation productivity over the supply and demand drought indices in the entire territory.

The variation in vegetation productivity appears to be associated with climate variation rather than anthropogenic factors (e.g., an increase in water demand by irrigated crops). Even though switching to more demanding crops on the land could increase the impact of drought on vegetation, this would need to be more thoroughly investigated, for instance at the watershed level.

The results of this study could help in the development of a robust forecasting system for land cover classes in Chile, helping to improve preparedness for climate change impacts on vegetation.

>>>>

---
title: The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile
#subtitle: A Short Subtitle
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.cl
    affiliations: 
        - id: umayor
          name: Universidad Mayor
          department: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad
          #address: Street Address
          city: Santiago, Chile
          #state: State
          postal-code: 7500994
        - id: odes
          name: Universidad Mayor
          department: Observatorio de Sequía para la Agricultura y la Biodiversidad de Chile (ODES)
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Anton Vrieling
    email: a.vrieling@utwente.nl
    affiliations:
        - id: itc
          name: University of Twente
          department: Faculty of Geo-Information Science and Earth Observation (ITC)
          #address: Street Address
          city:  Enschede, The Netherlands
  - name: Francisco Meza
  - name: Iongel Duran-Llacer
  - name: Francisco Fernández
    email: francisco.fernandez@uss.cl
    affiliations:
        - id: uss
          name: Universidad San Sebastian
          department: 
          #address: Street Address
          city:  
  - name: Alejandro Venegas-González
  - name: Nicolas Raab
  - name: Dylan Craven        
  #   note: |
  #     Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  # - name: Cat Memes
  #   email: cat@example.com
  #   affiliations:
  #       - ref: another-u
  #   note: Yet another author footnote.
  # - name: Derek Zoolander
  #   email: derek@example.com
  #   affilations:
  #       - ref: some-tech
abstract: |
 The north-central region of Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a mega-drought. The aim of our study is to evaluate the interaction of drought, land cover change, and vegetation productivity over continental Chile. We used drought indices for atmospheric evaporative demand (AED), water supply, and soil moisture from short-term (1, 3, 6 months) to long-term (12, 24, 36 months) time scales. We derived the drought indices using monthly ERA5-Land reanalysis data spanning from 1981 to 2023. We utilized the Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive information on annual land cover and monthly vegetation productivity. Our results showed that from south to north, Chile has a declining trend in water supply, and across the whole country, there is an increase in AED. These trends are stronger at longer time scales. The trend in vegetation productivity in the north-central area is affecting, to a higher degree, shrubland and savanna, followed by croplands and forests. The drought explains about 30% of the change in land cover type across Chile for forest, grassland, shrubland, and savanna. The increase in AED is the main driver of the change in land cover, followed by a reduction in precipitation and soil moisture. The change in vegetation productivity has been severe in the north-central part of the country for all land cover types, particularly savanna, shrubland, and croplands. The anomaly in soil moisture over the past 12 months (SSI-12) is the main variable explaining these changes, followed by anomalies in cumulated precipitation over one to two years (SPI-12 and SPI-24). Our results provide insightful information that would help in developing adaptation measures for ecosystems in Chile to cope with climate change and drought.

keywords: 
  - drought
  - land cover change
  - vegetation productivity
  - satellite
date: last-modified
bibliography: references.bib
keep-tex: false
include-before-body:
  - text: |
      \captionsetup{justification=raggedright,singlelinecheck=false}
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      layout: onecolumn
      model: 3p
      cite-style: authoryear
      natbiboptions: numbers
    include-in-header: 
      text: \usepackage{lineno}\linenumbers
            \usepackage{multirow}
            \usepackage{lscape}
            \newcommand{\blandscape}{\begin{landscape}}
            \newcommand{\elandscape}{\end{landscape}}
  elsevier-docx:
    keep-tex: true
    number-sections: true
    journal:
      name: Journal Name
      formatting: preprint
      layout: onecolumn
      model: 3p
      cite-style: authoryear
      natbiboptions: numbers
    include-in-header: 
      text: \usepackage{lineno}\linenumbers
            \usepackage{multirow}
            \usepackage{lscape}
            \newcommand{\blandscape}{\begin{landscape}}
            \newcommand{\elandscape}{\end{landscape}}
---

@@

For the purpose of monitoring water supply drought, we used the well-known Standardized Precipitation Index (SPI), which relies on precipitation data. To evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI), developed by @Hobbins2016 and @McEvoy2016, which is based on the AED. The United States currently monitors drought using the EDDI (https://psl.noaa.gov/eddi/) as an experimental index. To consider the combined effect of water supply and demand, we selected the SPEI [@Vicente-Serrano2010]. For SPEI, an auxiliary variable $D = P-AED$ is calculated. Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions [@Li2022]. Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) [@Hao2013;@AghaKouchak2014]. In our case, for the SSI, we used the average soil moisture from ERA5L at 1m depth. Finally, for the proxy of productivity, we used the zcNDVI proposed by @Zambrano2018, which was derived from the monthly time series of NDVI derived from MOD13A1. All the indices are multi-scalar and can be used for the analysis of short- to long-term droughts.

@@

The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 1981–2023 in order to be used for short- to long-term evaluation of drought. In the case of the proxy of vegetation productivity (zcNDVI), it was calculated for a time scale of six months at monthly frequency for 2000–2023. For zcNDVI, we test time scales of 1, 3, 6, and 12 months in December and their correlation with net primary production (NPP) obtained from the MOD17A3HGF product from MODIS. We choose to use six months because r-squared with NPP increases from one to six months and from six to 12 months has little improvement. The r-squared were between 0.31 for forest and 0.72 for shrubland (see supplementary material in Section S5).

@@

## Drought information to aid in adaptation

Different climate components, such as evaporative water demand, water supply, soil moisture, and their impact on vegetation, should be considered when evaluating the multi-dimensional nature of drought. For a better understanding of the propagation of drought [@VanLoon2012] from meteorological to hydrological drought, we should consider the climatic response at different time scales, ranging from short to long. Furthermore, we must make this information publicly available and update it frequently to aid in the development of adaptation policies. The drought observatory for agriculture and biodiversity of Chile (ODES) shares this information aggregated for multiple administrative and hydrological units across Chile ([https://odes-chile.org/app/unidades](https://odes-chile.org/app/unidades)), with the goal of helping to prepare for future climatic conditions.

@@

There is a trend toward decreasing water supply (SPI, SPEI, and SSI) in most of Chile; just in the southern part, there is an increase. The trend is most strong in the north-central zone. The whole country showed an increase in water demand (AED). The magnitude of the trend increases over longer time scales, which is evidence that the deficit is impacting the hydrological system in Chile. The trend in vegetation productivity in the north-central area is affecting, to a higher degree, shrubland and savanna, followed by croplands and forests. The most important changes in land cover are the increase of forest in "Sur" and shrubland and grassland in "Centro;" and the increase of savanna in "Centro" and "Sur.".

The drought explains about 30% of the change in land cover type across Chile for forest, grassland, shrubland, and savanna. There is no evidence of an effect of drought on the change in cropland surface area. The increase in AED is the main driver of the change in land cover, followed by a reduction in precipitation and soil moisture. The drought time scales vary regarding the land cover type.

The change in vegetation productivity has been severe in the north-central part of the country for all land cover types, particularly savanna, shrubland, and croplands. The anomaly in soil moisture over the past 12 months is the main variable explaining these changes, followed by anomalies in cumulated precipitation over one to two years. The variation in AED seems to intensify the drought impact on vegetation productivity.

The results of this study provide insightful information that would help in developing adaptation measures for ecosystems in Chile to cope with climate change and drought.

# Acknowledgment

The National Research and Development Agency of Chile (ANID) funded this study through the drought emergency project FSEQ210022, Fondecyt Iniciación N°11190360, and Fondecyt Regular N°1210526.

### 2024-04-04 article/paper_drought_Chile2023.tex
table landcover

\begin{frontmatter}
\title{The effects of multi-dimensional drought on land cover change and
vegetation productivity in continental Chile}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl} 
\author[2]{Anton Vrieling%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[]{Francisco Meza%
%
}

@@

\author[3]{Francisco Fernández%
%
}
 \ead{francisco.fernandez@uss.cl} 
\author[]{Alejandro Venegas-González%
%
}

@@

\affiliation[1]{organization={Universidad Mayor, Hémera Centro de
Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en
Medio Ambiente y Sustentabilidad},city={Santiago,
Chile},postcode={7500994},postcodesep={}}
\affiliation[2]{organization={University of Twente, Faculty of
Geo-Information Science and Earth Observation (ITC)},city={Enschede, The
Netherlands},postcodesep={}}
\affiliation[3]{organization={Universidad San
Sebastian},,postcodesep={}}

@@

\begin{abstract}
The north-central region of Chile has been the focus of research studies
due to the persistent decrease in water supply, which is impacting the
hydrological system and vegetation development. This persistent period
of water scarcity has been defined as a megadrought. The aim of our
study is to evaluate the land cover change over continental Chile and to
examine how this is connected to drought indices of water supply,
atmospheric evaporative demand (AED), soil moisture, and their effects
on vegetation productivity. The drought indices were derived using
monthly ERA5-Land reanalysis data spanning from 1981 to 2023. The
Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets were
utilized to obtain information on annual land cover and monthly
vegetation productivity. We analyzed short- (1, 3, 6 months) to
long-term (12, 24, 36 months) time scales of drought. Our results showed
that land cover change was highest in the south-central part of the
country, reaching changes as high as 36\% in the surface type. The water
demand has increased for the whole country, with a major increase in the
north. The AED and soil moisture evidence a decreasing trend, which
decreases at longer time scales and from north to south. The extreme
south part of the country shows an increase in supply. Vegetation
productivity has a negative trend in the north-central region for all
land cover types. On the other hand, forests seem to be the most
resistant type to drought. The types that show to be most affected by
variation in climate conditions are shrublands, savannas, and croplands.
The drought indices that have the capability of explaining to a major
degree the variance in vegetation productivity are the ones that
consider soil moisture for twelve months and the combined effect of
precipitation and AED for 24 and 12 months. The results indicate that
the north-central region is the most sensitive to water supply deficits
lasting longer than a year.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

For the purpose of monitoring water supply drought, we used the
well-known Standardized Precipitation Index (SPI), which relies on
precipitation data. To evaluate water demand, we chose the Evaporative
Demand Drought Index (EDDI), developed by \citet{Hobbins2016} and
\citet{McEvoy2016}, which is based on the AED. The United States
currently monitors drought using the EDDI (https://psl.noaa.gov/eddi/)
as an experimental index. To consider the combined effect of water
supply and demand, we selected the SPEI \citep{Vicente-Serrano2010}. For
SPEI, an auxiliary variable \(D = P-AED\) is calculated. Soil moisture
is the main driver of vegetation productivity, particularly in semi-arid
regions \citep{Li2022}. Hence, for soil water drought, we used the SSI
(Standardized Soil Moisture Index) \citep{Hao2013, AghaKouchak2014}. In
our case, for the SSI, we used the average soil moisture from ERA5L at
1m depth. Finally, for the proxy of productivity, we used the zcNDVI
proposed by \citet{Zambrano2018}, which was derived from the monthly
time series of NDVI retrieved from MOD13A1. All the indices are
multi-scalar and can be used for the analysis of short- to long-term
droughts.

@@

The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
and 36 months at a monthly frequency for 1981--2023 in order to be used
for short- to long-term evaluation of drought. In the case of the proxy
of vegetation productivity (zcNDVI), it was calculated for a time scale
of six months at monthly frequency for 2000--2023. For zcNDVI, we test
time scales of 1, 3, 6, and 12 months in December and its correlation
with net primary production (NPP) obtained from the MOD17A3HGF product
from MODIS. We choose to use six months because r-squared with NPP
increases from one to six months and from six to 12 months has little
improvement (see supplementary material in Section S5).

@@

\hypertarget{future-outlook-to-complete}{%
\subsection{Future outlook (to
complete)}\label{future-outlook-to-complete}}

@@

There is a trend toward decreasing water supply in most parts of Chile,
particularly in the ``Centro'' and ``Norte Chico'' regions. The whole
country showed an increase in AED. Vegetation productivity only showed a
decrease in the ``Norte Chico'' and ``Centro,'' being highest for
shrubland and croplands. Forest is the land cover most resistant to
drought, as shown along Chile, and shrubland and cropland are the most
sensitive.

A soil moisture deficit of 12 months (SSI-12) is highly correlated with
vegetation productivity for the land cover classes of shrubland,
savannas, croplands, and forest in ``Norte Chico'' and ``Centro.'' For
the southern part of the country with humid conditions, the correlation
with SSI decreases. Soil moisture overcomes the capacity to explain
vegetation productivity over the supply and demand drought indices in
the entire territory.

The variation in vegetation productivity appears to be associated with
climate variation rather than anthropogenic factors (e.g., an increase
in water demand by irrigated crops). Even though switching to more
demanding crops on the land could increase the impact of drought on
vegetation, this would need to be more thoroughly investigated, for
instance at the watershed level.

The results of this study could help in the development of a robust
forecasting system for land cover classes in Chile, helping to improve
preparedness for climate change impacts on vegetation.

>>>>

\begin{frontmatter}
\title{The effects of multi-dimensional drought on land cover change and
vegetation productivity in continental Chile}
\author[1,2]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl} 
\author[3]{Anton Vrieling%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[]{Francisco Meza%
%
}

@@

\author[4]{Francisco Fernández%
%
}
 \ead{francisco.fernandez@uss.cl} 
\author[]{Alejandro Venegas-González%
%
}

@@

\affiliation[1]{organization={Universidad Mayor, Hémera Centro de
Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en
Medio Ambiente y Sustentabilidad},city={Santiago,
Chile},postcode={7500994},postcodesep={}}
\affiliation[2]{organization={Universidad Mayor, Observatorio de Sequía
para la Agricultura y la Biodiversidad de Chile (ODES)},,postcodesep={}}
\affiliation[3]{organization={University of Twente, Faculty of
Geo-Information Science and Earth Observation (ITC)},city={Enschede, The
Netherlands},postcodesep={}}
\affiliation[4]{organization={Universidad San
Sebastian},,postcodesep={}}

@@

\begin{abstract}
The north-central region of Chile has been the focus of research studies
due to the persistent decrease in water supply, which is impacting the
hydrological system and vegetation development. This persistent period
of water scarcity has been defined as a mega-drought. The aim of our
study is to evaluate the interaction of drought, land cover change, and
vegetation productivity over continental Chile. We used drought indices
for atmospheric evaporative demand (AED), water supply, and soil
moisture from short-term (1, 3, 6 months) to long-term (12, 24, 36
months) time scales. We derived the drought indices using monthly
ERA5-Land reanalysis data spanning from 1981 to 2023. We utilized the
Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive
information on annual land cover and monthly vegetation productivity.
Our results showed that from south to north, Chile has a declining trend
in water supply, and across the whole country, there is an increase in
AED. These trends are stronger at longer time scales. The trend in
vegetation productivity in the north-central area is affecting, to a
higher degree, shrubland and savanna, followed by croplands and forests.
The drought explains about 30\% of the change in land cover type across
Chile for forest, grassland, shrubland, and savanna. The increase in AED
is the main driver of the change in land cover, followed by a reduction
in precipitation and soil moisture. The change in vegetation
productivity has been severe in the north-central part of the country
for all land cover types, particularly savanna, shrubland, and
croplands. The anomaly in soil moisture over the past 12 months (SSI-12)
is the main variable explaining these changes, followed by anomalies in
cumulated precipitation over one to two years (SPI-12 and SPI-24). Our
results provide insightful information that would help in developing
adaptation measures for ecosystems in Chile to cope with climate change
and drought.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

For the purpose of monitoring water supply drought, we used the
well-known Standardized Precipitation Index (SPI), which relies on
precipitation data. To evaluate water demand, we chose the Evaporative
Demand Drought Index (EDDI), developed by \citet{Hobbins2016} and
\citet{McEvoy2016}, which is based on the AED. The United States
currently monitors drought using the EDDI (https://psl.noaa.gov/eddi/)
as an experimental index. To consider the combined effect of water
supply and demand, we selected the SPEI \citep{Vicente-Serrano2010}. For
SPEI, an auxiliary variable \(D = P-AED\) is calculated. Soil moisture
is the main driver of vegetation productivity, particularly in semi-arid
regions \citep{Li2022}. Hence, for soil water drought, we used the SSI
(Standardized Soil Moisture Index) \citep{Hao2013, AghaKouchak2014}. In
our case, for the SSI, we used the average soil moisture from ERA5L at
1m depth. Finally, for the proxy of productivity, we used the zcNDVI
proposed by \citet{Zambrano2018}, which was derived from the monthly
time series of NDVI derived from MOD13A1. All the indices are
multi-scalar and can be used for the analysis of short- to long-term
droughts.

@@

The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
and 36 months at a monthly frequency for 1981--2023 in order to be used
for short- to long-term evaluation of drought. In the case of the proxy
of vegetation productivity (zcNDVI), it was calculated for a time scale
of six months at monthly frequency for 2000--2023. For zcNDVI, we test
time scales of 1, 3, 6, and 12 months in December and their correlation
with net primary production (NPP) obtained from the MOD17A3HGF product
from MODIS. We choose to use six months because r-squared with NPP
increases from one to six months and from six to 12 months has little
improvement. The r-squared were between 0.31 for forest and 0.72 for
shrubland (see supplementary material in Section S5).

@@

\hypertarget{drought-information-to-aid-in-adaptation}{%
\subsection{Drought information to aid in
adaptation}\label{drought-information-to-aid-in-adaptation}}

Different climate components, such as evaporative water demand, water
supply, soil moisture, and their impact on vegetation, should be
considered when evaluating the multi-dimensional nature of drought. For
a better understanding of the propagation of drought \citep{VanLoon2012}
from meteorological to hydrological drought, we should consider the
climatic response at different time scales, ranging from short to long.
Furthermore, we must make this information publicly available and update
it frequently to aid in the development of adaptation policies. The
drought observatory for agriculture and biodiversity of Chile (ODES)
shares this information aggregated for multiple administrative and
hydrological units across Chile
(\url{https://odes-chile.org/app/unidades}), with the goal of helping to
prepare for future climatic conditions.

@@

There is a trend toward decreasing water supply (SPI, SPEI, and SSI) in
most of Chile; just in the southern part, there is an increase. The
trend is most strong in the north-central zone. The whole country showed
an increase in water demand (AED). The magnitude of the trend increases
over longer time scales, which is evidence that the deficit is impacting
the hydrological system in Chile. The trend in vegetation productivity
in the north-central area is affecting, to a higher degree, shrubland
and savanna, followed by croplands and forests. The most important
changes in land cover are the increase of forest in ``Sur'' and
shrubland and grassland in ``Centro;'' and the increase of savanna in
``Centro'' and ``Sur.''.

The drought explains about 30\% of the change in land cover type across
Chile for forest, grassland, shrubland, and savanna. There is no
evidence of an effect of drought on the change in cropland surface area.
The increase in AED is the main driver of the change in land cover,
followed by a reduction in precipitation and soil moisture. The drought
time scales vary regarding the land cover type.

The change in vegetation productivity has been severe in the
north-central part of the country for all land cover types, particularly
savanna, shrubland, and croplands. The anomaly in soil moisture over the
past 12 months is the main variable explaining these changes, followed
by anomalies in cumulated precipitation over one to two years. The
variation in AED seems to intensify the drought impact on vegetation
productivity.

The results of this study provide insightful information that would help
in developing adaptation measures for ecosystems in Chile to cope with
climate change and drought.

\hypertarget{acknowledgment}{%
\section{Acknowledgment}\label{acknowledgment}}

The National Research and Development Agency of Chile (ANID) funded this
study through the drought emergency project FSEQ210022, Fondecyt
Iniciación N°11190360, and Fondecyt Regular N°1210526.
